Designing Types for R, Empirically

### 2020-05-15 paper/main.tex
minor
4.2 wip

\paragraph{Raw types}
The type information output by the tool includes the \emph{type tag} of each
value. The R internal types are translated to names in the proposed type
system. The next bit of information output is the \emph{class} which is an
optional list of names, it may be absent, and, in some cases, it may be
implicit (i.e. the interpreter blesses some values with the \k{matrix} and
\k{array} classes even when they have no attributes).

Depending on the value's type tag, the tool collects further information:

@@

The type information we collect is far more detailed than the types we
actually generate (which we discuss in Section~\ref{sec:types}), and so the
next stage of our inference process is to distill the rich type information
into simpler, manageable types.  For each unique observed call trace, we
translate the representation of the type of each of the function arguments
and returns according to the design of our type system: for instance, we
might have determined that a function returned a vector of integer with 4
elements, with names, that was free of NA values, in which case we would
write out \code{integer[]} as the type.

Once types have been simplified in this way, for each function we move to
consolidate its multiple call traces into a single function type.  We do
this by combining the observed types of each argument into one type
according to subtyping rules that are part of our design.  The precise
subtyping rules we used will be discussed further in
Section~\ref{sec:typesystemdesign}.  If we cannot consolidate multiple
observed argument types into a single type, we simply union the disparate
types together.

While our analysis collects more detailed information than necessary for
type generation, the information proved useful for building up a usage
profile of the language.  This profile informed our language design: based
on the rich type information we collected, we were able to determine types
that would be useful for capturing the observed behaviour.  We discuss this
iterative, data-driven language design process in Section~\ref{types},
highlighting the reasoning behind our type system design.

Thus far, we have discussed \typetracer and our strategies to take the data
from the dynamic analysis and generate function types.  This work also
presents a framework for specifying these types in R code, as well as
generating contract checks for these types.  We dub this tool \contractr,
discussed next.

%
%
%
\subsection{Checking types signatures with contracts}

@@

\AT{Filip}

%
%
%
%
%
%
\section{Related Work}
\label{sec:relatedwork}

\AT{Good to mention the related projects, e.g. how this was undertaken in
  other languages, like Ruby.  There's a lot of literature in that space I
  think, I've read some of it but not all.  We also have some related work
  in the old paper, I'll grab it and put it here.}

%
%
%
%
%
%
\section{Conclusion}

@@

%
%
%
%
\subsection{Threats to Validity}

%
%
%
%
\subsection{Future Work}

%
%
\subsubsection{Object Orientation in R}

%
%
\subsubsection{Types for Data Frames}

% ======================================
%
% End
%
%

\bibliography{bib/biblio,bib/jv,bib/r,bib/new,bib/gradual}

\end{document}
\appendix
\section{Appendix}

>>>>

\paragraph{Raw types}
The type information output by the tool includes the \emph{type tag} of each
value. The R internal types are translated to names in the proposed type
system. The next bit of information output is the \emph{class} which is an
optional list of names, it may be absent, and, in some cases, it may be
implicit (i.e. the interpreter blesses some values with the \k{matrix} and
\k{array} classes even when they have no attributes). 
Depending on the value's type tag, the tool collects further information:

@@

The output of the \typetracer tool consists of a set of tuples of row types,
each representing a function call in some program's execution. The inference
step consolidate the different tuples corresponding to a particular function
definition across multiple programs and distills them into a single type
signature.  The shape of the infered function signatures is:

\medskip
\FUN{ T_{1,1} \k{~|~} T_{1,i}, \dots, T_{n,1} \k{~|~} T_{n,j}}{T_1\k{~|~} T_k}
\medskip

In other words, we take the union of the types occurring at individual
argument positions rather than an union of function types. Furthermore we
apply some transformation on the types to keep the size of types in check.

\begin{tabular}{rclll}
  \T \k{~|~} \T  & $\Rightarrow$ &  \T &  \\
  \T \k{~|~} $T'$ & $\Rightarrow$ &  \T & \it{iff} $T' <: T$  \\
   $\null \k{~|~} \VEC{\S_1} \k{~|~}\dots \VEC{\S_n}$  & $\Rightarrow$ &
    $ \NAVEC{\S_1} \k{~|~}\dots \NAVEC{\S_n}$ &  \\  
\end{tabular}

Assuming that type sequences can be reordered freely, redundant types can be
deleted, types that are subsumed by subtyping can be removed, and nullable
types can be introduced to replace \null.

\subsection{Checking types signatures with contracts}

@@

\section{Conclusion}

@@

\bibliography{bib/biblio,bib/jv,bib/r,bib/new,bib/gradual}
\end{document}
